---
title: "Starting with with the dplyr, ggplot2 packages, and the pipe operator |>"
author: "Han Olff"
date: "2024-08-29"
format: html
editor: 
  markdown: 
    wrap: 72
    editor_options: 
    chunk_output_type: console
---

# Working with the dplyr and ggplot2 packages

Load the required libraries :

```{r message = FALSE, warning = FALSE}
# restore your library (keep this at the start of every script)
renv::restore()

# clear the R environment
rm(list = ls())

# load the required packages
library(tidyverse)  # this loads a group of related packages, including readr, dplyr and ggplot2
```

We will work with an online database of the Schiermonnikoog transect
study, that you will collect additional data for next week in the field.
In this example, we work with measurements of cockles (a bivalve
mollusc) on their width and length. From the database Epibenthos, we
will work with the table
[FactCockles](https://docs.google.com/spreadsheets/d/1E1vUWAHhse7fhjBf94Kiog3Rsj7IkMuFtP-BTQX2oI8/edit?gid=1538766002#gid=1538766002).
See the documentation of the different variables in the table
[MetVariables](https://docs.google.com/spreadsheets/d/1E1vUWAHhse7fhjBf94Kiog3Rsj7IkMuFtP-BTQX2oI8/edit?gid=1290622213#gid=1290622213)

### Read the datafile from the Google sheets database

To read the data in R, you first need to know the published csv link of
the FactCockles table. In the database, you can find this link in the
table MetTables. It is produced in Google Sheets throught the menu
File/Share/Publish to web and then selecting the table and output as
csv.

Read the FactCockleSize table with read_csv, which reads it as tibble
(formatted dataframe)

```{r}
FactCockleSize <- readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSpormjGiM4uWMTIZYm9Upt5j1Ige_Wu9CrhGGdvjXwxmHP2R6S7JDyEmiGTn3fXihl5s5yBn_sdo8h/pub?gid=1538766002&single=true&output=csv")
print(FactCockleSize)
names(FactCockleSize)
nrow(FactCockleSize)
```

Plot with ggplot the relation between cockle thickness (thickness_mm, as
x) and length (length_mm), showing each year with a different color, and
add a regression line through all the years

```{r}
ggplot2::ggplot(data=FactCockleSize,    # what data table to use
                mapping=aes(x=length_mm,y=thickness_mm)) + # what is on the axes?
               geom_point()

# there is a big outlier, likely a data entery mistake, remove it 
FactCockleSize |> dplyr::filter(thickness_mm<500)

FactCockleSize <- readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSpormjGiM4uWMTIZYm9Upt5j1Ige_Wu9CrhGGdvjXwxmHP2R6S7JDyEmiGTn3fXihl5s5yBn_sdo8h/pub?gid=1538766002&single=true&output=csv") |> 
  dplyr::filter(CockleObs_ID!=1531, CockleObs_ID!=469) |>
  dplyr::mutate(year=factor(year))  # make a year a factor
  # adding this comment is to get the mistake or the outlier out of the tibble in this case cockle 1531 and 469
  # mutate function is for when you want R to treat the variable as categories or as a group. Mainly when it is numerical you need to use this function. When it is text is does not necessarily be the case. Mutate function can also be used for different things, look it up :)
print(FactCockleSize)

# plot the data
ggplot2::ggplot(data=FactCockleSize,    # what data table to use
                mapping=aes(x=length_mm,y=thickness_mm)) + # what is on the axes?
               geom_point()

# find that other outlier
FactCockleSize |> dplyr::filter(thickness_mm>10 & length_mm<5)
# also remove 469 above 
```

Further explore the plot with a regression line through all the data

```{r}
# plot the data with a regression line
ggplot2::ggplot(data=FactCockleSize,    # what data table to use
                mapping=aes(x=length_mm,y=thickness_mm)) + # what is on the axes?
         geom_point() +
         geom_smooth(method="lm")
# calculate a linear regression
model_lm<-lm(thickness_mm~length_mm, data=FactCockleSize)   # ~ = y is explained by x
summary(model_lm)

# equation: thickness_mm=-0.353442+0.699362*length_mm
# a cockle of length 20 mm has an predicted width of: 
-0.353442+0.699362*20
# this is regression is significant, R2=0.90 (90% of the variation of cockle length is explained in the graph(?)), P<0.001 (***)
```

Make same plot but showing a separate regression line per year

```{r}
# color the points by year, but plot one regression line
ggplot2::ggplot(data=FactCockleSize,    # what data table to use
                mapping=aes(x=length_mm,y=thickness_mm)) + # what is on the axes?
         geom_point(mapping=aes(col=year)) +
         geom_smooth(method="lm")
# different mapping statement for different geom stations, but not always necessary, for instance, in this case there is no mapping for geom_smooth
```

Make a panel plot where with each year is shown as a separate graph

```{r}
drop_na(FactCockleSize) |> # what data table to use 
      ggplot2::ggplot(mapping=aes(x=length_mm,y=thickness_mm)) + # what is on the axes?
         geom_point(mapping=aes(col=year)) +
         geom_smooth(method="lm") +
         facet_wrap(~year)  
# facet_wrap is to make a panel plot, by using ~ you specify in which way the graphs have to be separate
# to filter the data set you can delete the data= out of the ggplot and pipe it in with |>. Since we want to filter out certain years, we use drop_na() to only get graphs from the years with data. 
```

We conclude from this that: \* there were two important outliers in the
dataset that were removed after visual inspection \* the regression
between length and width is about the same for every year, we safely can
use only length as a proxy for the biomass of an individual cockle
